### 2020-11-12 latex/sections/background.tex
latexing

In this chapter, we discuss traditional artificial intelligence in the context of search the algorithms we used, what some other similar methods exist as well as the field in general. We allocate a large portion of the section to Breakthrough as that will be the testbed for future sections

@@

When researching Artificial Intelligence it is important to select an environment that is a suitable abstraction for the task at hand. Environments vary significantly and can be identified by their characteristic. The characteristics that are generally used to describe environments can be seen in Table~\ref{tab:env_characteristics}. \cite{Russell:AIModern}

\begin{table}[h]
  \centering
  \begin{tabular}{|c|c|p{6cm}|}
    \hline
    \textbf{characteristic} & \textbf{Values}              & \textbf{Description}                                                \\
    \hline
    Observable              & Fully, Partially             & How much of the environment can your agent percieve.                \\
    \hline
    Agents                  & Single, Multi                & Are there multiple agents playing in the environment.               \\
    \hline
    Deterministic           & Deterministic, Stochastic    & Do the actions your agent do deterministicly impact the environment \\
    \hline
    Episodic                & Sequential, Episodic         & Are actions episodic or sequential                                  \\
    \hline
    Static                  & Static, Semi-Static, Dynamic & ehh                                                                 \\
    \hline
    Discrete                & Discrete, Continuous         & Is your environment discrete w.r.t actions (does it end)            \\
    \hline
  \end{tabular}
  \caption{Characteristics of environments}
  \label{tab:env_characteristics}
\end{table}

Categorizing environments like this gives you the power to find an environment in which a method works and know it can be applied to different environments with the same characteristics. Talk about agents in the context of environments as the entities that act within the environment. A game player is an agent as well as the neural network model that exists within a car that can drive in the real world.

@@

The game breakthrough is a simplified version of chess, the game is set up on a $MxN$ board with cells like in chess, and each player starts with two rows of pawns at either end. The objective of the game is to move one of the pawn pieces to the opposite end of the board. A player wins if either they have reached the opposite end of the board or have captured all of his opponents' pawns. The pawns differ from chess pawns in such a way that they can not move two squares on the first move and, they can move diagonally as well as forward, this leads to the game being impossible to draw as pieces are always able to move. An example of an initial board in breakthrough can be seen in Figure~\ref{fig:initbtboard}

@@

  \caption{Example breakthrough board}
  \label{fig:initbtboard}
\end{figure}

\begin{table}[h]
  \centering
  \begin{tabular}{|c|c|}\hline
    \textbf{Characteristic} & \textbf{Value} \\\hline
    Observable              & Fully          \\
    Agents                  & Multi          \\
    Deterministic           & Deterministic  \\
    Episodic                & Sequential     \\
    Static                  & Static         \\
    Discrete                & Discrete       \\\hline
  \end{tabular}
  \caption{Categorization of Breakthrough}
  \label{tab:breakthrough_cat}
\end{table}

Categorizing Breakthrough with the characteristics described in Section \ref{sec:environments}. We end up with the description
shown in Table \ref{tab:breakthrough_cat}. These characteristics are identical to that of Tic-Tac-Toe, and Chess.
This categorization is the most common in board games where two player compete.

@@

To evaluate the game of Breakthrough we can consider many heuristics (higher-level concepts) for instance a very simple heuristic would be the number of pieces each player has left. This heuristics gives us some insight into how well the game is progressing, but obviously, there are cases where this doesn't tell us much, as in cases when your opponent has a single piece left that is on the row immediately before the row needed for him to win. No matter how many pieces you have left, this state is bad for you if you're not able to capture that piece. An example of such a state can be seen in Figure~\ref{fig:bt_h1_bad}.

\begin{figure}[]
  \centering
  \breakthrough{8/8/pppp4/pp3P2/2p5/8/8/PPP5 w - - 0 1}
  \caption{Breakthrough board where Heuristic 1 doesn't work well}
  \label{fig:bt_h1_bad}
\end{figure}

A different heuristic would be the distance of your most advanced pawn minus your opponents most advanced pawn, this heuristic could give you insight into how close you are to winning the game or how close your opponent is. As a higher-level concept, we can call this concept your aggressiveness, as it closely resembles how aggressive you are going for in for the win. Generally, in Breakthrough, it is favorable to move your whole team as a unit and play more defensively, so this heuristic is probably not one that will lead to the most wins.

@@

Traditionally methods for playing games were searching through the environment using a heuristic to guide the search. A simple way of doing a heuristic-based search would be to give all states a $0$ as a score and give states where you win a positive score. We say that that search algorithm is not guided, and the algorithm will probably have to evaluate every state in the state-space. This method of searching is generally extremely inefficient as the state-spaces of game environments are often extremely large, and sometimes infinite. For instance, an estimate of the state-space for breakthrough is $M * N * 2^{N*4}$ where $M$ is the height of the board $N$ is the width of the board. The $2^{N*4}$ represents each piece either being alive or captured. So for a small board, $5x4$ the upper-bound estimation of the state-space is $1310720$ states.

This is why a good heuristic is very valuable because we can disregard all states $S'$ that result from doing a move $a$ in state $S$ as they will only lead to worse outcomes.

@@

More modernly, these search methods have been amplified by Machine Learning, in such a way that we do not need to figure out a good heuristic for a given state, but rather, we apply a machine learning model to learn a function that takes in a state and returns an evaluation of that state. This can lead to a significant time reduction as we do not need to simulate a whole game from a state to receive its evaluation we rather receive the evaluation from the model.

@@

In the algorithm Monte-Carlo Tree Search, there is an agent within some environment. Where each node in the environment represents a state-action pair of the environment, by state-action pair what is meant it is some state and the action that brought the agent to that state. This pair should be unique within the environment. We will discuss these concepts within game-environments and therefore we might say player instead of agent and game instead of the environment.

@@

In MCTS there is a tree representing the game-environment. This tree consists of nodes $n_i$ where $i$ represents the point in time of the node, for example, $N_0$ in chess is the initial position and $N_x$ is some position in the middle of the game and $N_e$ is one of the states representing a position where there is either a draw or one player has won the match. Each of the nodes has $4$ values, $s$, $a$, $Q$, and $N$. These values represent these items, $s$ is the state of the environment, $a$ is the action that brought the previous node $n_(i-1)$ to node $n_i$, $Q$ is the average value of the node (value meaning the outcome of the game), and $N$ the number of times the MCTS algorithm has visited this node. The values $s$ and $a$ uniquely identify a position in the environment and are often called state-action pairs.

@@

\subsection*{Expansion}
Then the expansion phase expands the node generating all of $(s,a)$'s children $(s',a')$ are generated
and connected to $(s,a)$. This is done by applying all actions $a'$ to $(s,a)$. These children are the
product of applying each action $a'$ to $s$ in the parent node.

\subsection*{Rollout}
Next during rollout actions from $(s,a)$ are randomly selected to move to $(s',a')$, then repeated
to go to $(s'',a'')$, until a terminal node within the environment is reached. By terminal, we mean a
state in which the game is finished. A terminal node in MCTS can generally return any value, but in the
context of this paper, we only return (+1 white wins, -1 black wins, 0 draw).

@@

When training a neural network the UCT formula is modified slightly to prefer selecting nodes
that the neural network values highly by introducing a second scalar to the formula $f(s,a) = (p,v)$,
the resulting formula is described in Equation \ref{PUCT_formula}, and is called PUCT. Secondly, the backpropagation process
is modified to instead of doing rollout/simulation to receive a reward the predicted value $v$ from
the neural network is used instead.

\begin{equation} \label{PUCT_formula}
  \text{Child PUCT value} = \frac{Q_{(s',a')}}{N_{(s',a')}} + c_{uct} * P((s,a)) * \frac{\sqrt{log(N_{(s,a)})}}{N_{(s',a')}}
\end{equation}

@@

\subsubsection{Saliency Maps}
Within XAI many methods have been developed to try to evaluate
ANN's, these methods are often referred as model interpretability methods.
In the field image recognition there has been a lot of work examining which
pixels of an image the model deems important. Arguably the most popular
method for this is Saliency Maps\cite{Koch:saliency}. There the pixel values the model
deems important are colored in s.t. a human can examine the image and get a
sense of what portions of the image are important to the model, an example of
a classification of a dog can be seen in Figure \ref{fig:dog_saliency}.

@@

Methods for explaining models that aren't image recognition models include
shapley values, from the Lundberg \& Lee\cite{LundbergL:shapley}. There the input is examined against it's output, then iteratively
input values are selected to be fixed. Then the other input values are varied and
an average change in prediction is calculated. With this the shapley value can be
estimated for the fixed input value. This is done to examine which input values have
the strongest link to the output value. Shapley values on a dataset can give insights
on which input values the model deems important.

>>>>

In this chapter, we discuss traditional artificial intelligence in the context of search the algorithms we used, what some other similar methods exist as well as the field in general. We allocate a large portion of the section to Breakthrough as that will be the testbed for future sections. We discuss the different classes of machine learning as well as discuss neural networks in more depht.

@@

When researching Artificial Intelligence it is important to select an environment that is a suitable abstraction for the task at hand. Environments vary significantly and can be identified by their characteristic. The characteristics that are generally used to describe environments can be seen in Table~\ref{tab:env_characteristics}. This categorization of environments is described in the book Artificial Intelligence by Norvig \& Russell. \cite{Russell:AIModern}

\begin{table}[ht]
  \centering
  \begin{tabular}{|c|c|p{6cm}|}
    \hline
    \textbf{characteristic} & \textbf{Values}              & \textbf{Description}                                                                 \\
    \hline
    Observable              & Fully, Partially             & How much of the environment can your agent percieve.                                 \\
    \hline
    Agents                  & Single, Multi                & Are there multiple agents playing in the environment.                                \\
    \hline
    Deterministic           & Deterministic, Stochastic    & Do the actions your agent do deterministicly impact the environment.                 \\
    \hline
    Episodic                & Sequential, Episodic         & Are actions episodic or sequential.                                                  \\
    \hline
    Static                  & Static, Semi-Static, Dynamic & Does the environment without agent input, or does it wait until agents take actions. \\
    \hline
    Discrete                & Discrete, Continuous         & Is your environment discrete w.r.t actions.                                          \\
    \hline
  \end{tabular}
  \caption{Characteristics of environments}
  \label{tab:env_characteristics}
\end{table}

Categorizing environments like this gives you the power to find an environment in which a method works and know it can be applied to different environments with the same characteristics. Additionally, it allowes us to talk about agents in the context of environments as the entities that act within the environment.

@@

The game breakthrough is a simplified version of chess, the game is set up on a $MxN$ board with cells like in chess, and each player starts with two rows of pawns at either end. The objective of the game is for a player to move one of their pawns to the opposite end of the board. A player wins if either they have reached the opposite end of the board or have captured all of his opponents' pawns. The pawns differ from chess pawns in such a way that they can not move two squares on the first move and, they can move diagonally as well as forward, this leads to the game being impossible to draw as pieces are always able to move. An example of an initial board in breakthrough can be seen in Figure~\ref{fig:initbtboard}

@@

  \caption{Initial breakthrough board}
  \label{fig:initbtboard}
\end{figure}

\begin{table}[ht]
  \centering
  \begin{tabular}{|c|c|}\hline
    \textbf{Characteristic} & \textbf{Value} \\\hline
    Observable              & Fully          \\
    Agents                  & Multi          \\
    Deterministic           & Deterministic  \\
    Episodic                & Sequential     \\
    Static                  & Static         \\
    Discrete                & Discrete       \\\hline
  \end{tabular}
  \caption{Categorization of Breakthrough}
  \label{tab:breakthrough_cat}
\end{table}

@@

To evaluate the game of Breakthrough we can consider many heuristics (higher-level concepts) for instance a very simple heuristic would be a players Material Advantage. Material Advantage is the amount of pieces the player has minus the amount of pieces the opponent has. This heuristics gives us some insight into how well the game is progressing, but obviously, there are cases where this doesn't tell us much, as in cases when your opponent has a single piece left that is on the row immediately before the row needed for him to win. No matter how many pieces you have left, this state is bad for you if you're not able to capture that piece. An example of such a state can be seen in Figure~\ref{fig:bt_h1_bad}.

\begin{figure}[]
  \centering
  \breakthrough{8/8/pppp4/pp3P2/2p5/8/8/8 w - - 0 1}
  \caption{Breakthrough board where Material Advantage doesn't work well}
  \label{fig:bt_h1_bad}
\end{figure}

A different heuristic would be the distance of your most advanced pawn minus your opponents most advanced pawn, this heuristic could give you insight into how close you are to winning the game or how close your opponent is. As a higher-level concept, we can call this concept your aggressiveness, as it closely resembles how aggressive you are going for in for the win. Generally, in Breakthrough, it is favorable to move your whole team as a unit and play more defensively, and additional heuristics for Breakthrough will be discussed in detail later in this research.

@@

Traditionally methods for playing games were searching through the environment using a heuristic to guide the search. A simple way of doing a heuristic-based search would be to give all states a $0$ as a score and give states where you win a positive score. We say that that search algorithm is not guided, and the algorithm will probably have to evaluate every state in the state-space. This method of searching is generally extremely inefficient as the state-spaces of game environments are often extremely large, and sometimes infinite. For instance, an upper-bound estimate of the state-space for breakthrough is $3^{(M-1)*(N-1)}+2*N$ where $M$ is the height of the board $N$ is the width of the board. The $3^{(M-1)*(N-1)}$ represents each position of the board having either a white, or black piece, or being empty. So for a small board, $5x4$ the upper-bound estimation of the state-space is $531,449$ states.

This is why a good heuristic is very valuable because we can disregard all following states that result from doing a move some in a previous state as they will only lead to worse outcomes.

@@

More modernly, these search methods have been amplified by Machine Learning, in such a way that we do not need to figure out a good heuristic for a given state, but rather, we apply a machine learning model to learn a function that takes in a state and returns an evaluation of that state.\cite{neuralnetworksgames:michulke} This can lead to a significant time reduction as we do not need to simulate a whole game from a state to receive its evaluation we rather receive the evaluation from the model.

@@

In the algorithm Monte-Carlo Tree Search, there is an agent within some environment. Where each node in the environment represents a state-action pair of the environment, by state-action pair what is meant it is some state and the action that brought the agent to that state. This pair should be unique within the environment.

@@

In MCTS there is a tree representing the game-environment. This tree consists of nodes $n_i$ where $i$ represents the point in time of the node, for example, $N_0$ in chess is the initial position and $N_x$ is some position in the middle of the game and $N_e$ is one of the states representing a position where there is either a draw or one player has won the match. Each of the nodes has $4$ values, $s$, $a$, $Q$, and $N$. These values represent these items, $s$ is the state of the environment, $a$ is the action that brought the previous node $n_{i-1}$ to node $n_i$, $Q$ is the average reward from running the MCTS algorithm from this node, and $N$ the number of times the MCTS algorithm has visited this node. The values $s$ and $a$ uniquely identify a position in the environment and are often called state-action pairs.

@@

\subsection*{Expansion}
Then the expansion phase expands the node generating all of $(s,a)$'s children, $(s',a')$ are generated by applying all actions $a'$ in $(s,a)$.

\subsection*{Rollout}
Next during rollout, actions from $(s,a)$ are randomly selected to move to $(s',a')$, then repeated to go to $(s'',a'')$, until a terminal node within the environment is reached. By terminal, we mean a state in which the game is finished. A terminal node in MCTS can generally return any value, but in the context of this paper, we only return (+1 white wins, or -1 black wins).

@@

When training a neural network the UCT formula is modified slightly to prefer selecting nodes
that the neural network values highly by introducing a second scalar to the formula $f((s,a)) = (p,v)$, where $f$ is the neural network, $p$ is the policy vector returned by the neural network and $v$ is the predicted value from the neural network. The resulting formula is described in Equation \ref{PUCT_formula}, and is called PUCT. Secondly, the backpropagation process
is modified to instead of doing rollout/simulation to receive a reward the predicted value $v$ from the neural network is used instead.

\begin{equation} \label{PUCT_formula}
  \text{Child PUCT value} = \frac{Q_{(s',a')}}{N_{(s',a')}} + c_{uct} * p_{(s,a)} * \frac{\sqrt{log(N_{(s,a)})}}{N_{(s',a')}}
\end{equation}

@@

Model Interpretability is the more common approach of XAI, as it is

\subsubsection{Saliency Maps}

Within XAI many methods have been developed to try to evaluate
ANN's, these methods are often referred as model interpretability methods.
In the field image recognition there has been a lot of work examining which
pixels of an image the model deems important. Arguably the most popular
method for this is Saliency Maps\cite{Koch:saliency}. There the pixel values the model
deems important are colored in s.t. a human can examine the image and get a
sense of what portions of the image are important to the model, an example of
a classification of a dog can be seen in Figure \ref{fig:dog_saliency}.

@@

Methods for explaining models that aren't image recognition models include
Shapley values, from the Lundberg \& Lee\cite{LundbergL:shapley}. There the input is examined against it's output, then iteratively
input values are selected to be fixed. Then the other input values are varied and
an average change in prediction is calculated. With this the shapley value can be
estimated for the fixed input value. This is done to examine which input values have
the strongest link to the output value. Shapley values on a dataset can give insights
on which input values the model deems important.

### 2020-11-12 latex/sections/discussion.tex
latexing

\ifdraft{Here I will discuss the myriad of fields research like this can help
for instance for health organizations, patients, taxpayers, and pharma}
hey

@@

\ifdraft{summarize the work}
hey

@@

\ifdraft{conclude the work, discuss the significance of the work}
hey

>>>>

This section discusses the results and the future work that could span from this research. Additionally, we conclude the work.

@@

As seen in Chapter \ref{cha:results}, our trained agent does indeed quickly rise to using the higher-level concepts. The results are promising, as we see a popular concepts like Lorentz-Horey rise in emphasis for the neural network, and a simple but effective concept like Material Advantage rise slowly but not exessively. We find the Unity concept to be disappointing as stated earlier in this paper maintaining closeness of your pawns tends to be preferential. And lastly, for a the concept Aggressiveness, the fact that it falls off early is what we expected. It would have been the preferred result if we saw a greater variability in the concepts, but this could be a result from not enough training iterations, or not a complex enough model. Importantly, these concepts are only a few possibilities of an infinite set of concepts, and the neural network could be learning a completely different concept then those that we tested. However, that was not the focus of this research, it was to examine if a neural network does move towards HLC's that we humans use in games.

\section{Future Work}

To iterate on this research, training a neural network with greater computing power will allow the researchers to hopefully achieve a super-human neural network in Breakthrough. This would give a greater confidence in the learned concepts, possibly allowing for pedagological research on the topic regarding which concepts are optimal to train people in the game environment. This could also arise from a larger neural network architecture.

An avenue of research would be to use a trained AlphaZero model that plays chess to a super human ability, and be able to draw from a more diverse set of concepts. Furthermore, one could examine the play-style of the super-human model to extract concepts in order to construct new and improved heuristics for a non-neural network based agent.

Additionally applying this method to a greater set of environment would be an interesting field of research. We envision a self driving car agent being examined with respect to a higher level concept of aggressiveness, or a mortgage agent being examined with respect to a higher level concept of gender bias.

@@

The work done in this paper is only a first step in examining working agents in active environments with respect to human level concepts. If improved could have a great impact in our trust on neural network that improve our daily lives. While the testbed for the research was a discrete game environment, there is a clear way forward to a dynamic system with discrete human level concepts. Our results show that for an agent that clearly doesn't achieve human level performance it's actions are often selected with respect to human level concepts, up to 60\% of the actions.

### 2020-11-12 latex/sections/introduction.tex
latexing

\chapter{Introduction\label{cha:introduction}}
%% \ifdraft only shows the text in the first argument if you are in draft mode.
%% These directions will disappear in other modes.
\ifdraft{State the objectives of the exercise. Ask yourself:
  \underline{Why} did I design/create the item? What did I aim to
  achieve? What is the problem I am trying to solve?  How is my
  solution interesting or novel?}{}

The world of deep learning (DL) is exciting, we have models that can examine
images and very reliably be able to identify it's content. Deep learning models have
beaten Ophthalmologists in identifying diabetic retinopathy, they've identified
cancer cells where others have not. Deep learning models have even predicted
the likelihood a person will die in the following year with good accuracy (CITE).
These models have done these things extremely accurately, cheap, and fast.

@@

The field of examining deep learning models to gain a richer understanding
of how they work, and what makes them so much better than humans at various tasks
is still new. This is the field of Explainable Artificial Intelligence (XAI).
If we wish to continue using artificial intelligence (AI) and machine learning (ML),
to improve our lives we must examine how they work, this is not only to improve
the models themselves but could also augment our ability in many cases.
Furthermore, these explanations are a legal requirement now in many areas, namely, legal, and
medical. Recently, XAI has generally been focused on Model Interpretability (MI), in
particular, interpreting the model in such a way that we might predict what the model will
do given a particular input. Another class of XAI would be Model Explainability (ME), where we can adequately explain how the internal mechanics of the model works.
A simple example of when we're able to apply ME would be when we print out a decision tree
that is generated from an ML algorithm, with the decision tree nodes and is conditions we can see exactly what the tree will do with a given input. While in MI we would have to predict the results, without going into details as to why.

@@

We define the research questions

\begin{enumerate}
  \item Given a NN that plays Breakthrough very well, can we evaluate if that NN recognizes the 
  HLCs that human players use.
  \item Does a NN that trains itself using self-play learn these HLCs over time, and does it start to emphasize 
  the HLCs that are generally taught later to human players, more as it trains itself more.
  \item Does a NN that plays better than some humans focus on HLCs that are not generally 
  considered by those human players and it.
\end{enumerate}

>>>>

\chapter{Introduction\label{cha:introduction}}
%% \ifdraft only shows the text in the first argument if you are in draft mode.
%% These directions will disappear in other modes.

The world of deep learning (DL) is exciting, we have models that can examine
images and very reliably be able to identify it's content. Deep learning models have
beaten Ophthalmologists in identifying diabetic retinopathy, they've identified
cancer cells where others have not. Deep learning models are used for the bleeding
edge of protein folding, to gain further understanding of the underlying structure of organisms, such as, viruses. \cite{deepmind:alphafold}
These models have done these things extremely accurately, cheap, and fast.

@@

The field of examining deep learning models to gain a richer understanding
of how they work, and what makes them so much better than humans at various tasks
is still new. This is the field of Explainable Artificial Intelligence (XAI).
If we wish to continue using artificial intelligence (AI) and machine learning (ML),
to improve our lives we must examine how they work, this is not only to improve
the models themselves but could also augment our ability in many cases.
Furthermore, these explanations are a legal requirement now in many areas\cite{legalexplanation:goodman}, namely, legal, and
medical. Recently, XAI has generally been focused on Model Interpretability (MI), in
particular, interpreting the model in such a way that we might predict what the model will
do given a particular input. Another class of XAI would be Model Explainability (ME), where we can adequately explain how the internal mechanics of the model works.
A simple example of when we're able to apply ME would be when we print out a decision tree
that is generated from an ML algorithm, with the decision tree nodes and is conditions we can see exactly what the tree will do with a given input. While in MI we would have to predict the results, without going into details as to why.

@@

We define the research questions:

\begin{enumerate}
  \item Given a NN that plays Breakthrough very well, can we evaluate if that NN recognizes the
        HLCs that human players use.
  \item Does a NN that trains itself using self-play learn some HLCs over time, and does it start to emphasize
        the HLCs that are generally considered better, more as it trains more, and oppositely does it start to
        demphasize HLCs that are considered worse.
\end{enumerate}
